---
title: 'ST565: Time Series HW7'
author: "Tadesse Zemicheal"
output: pdf_document
layout: one-col

---

## Question 1

### 1. Derive the spectrum for an MA(1) process.

$$ \gamma(k) = \begin{cases}
\beta_1^2 \sigma^2 + \sigma^2 \text{\qquad for k = 0} \\
\beta_1 \sigma^2 \text{\qquad  for k = 1}\\
0 \text{\qquad for k >= 2}
\end{cases}
$$

$$ 
f(\omega) = 1/\pi \big[\gamma(0) + 2 \Sigma_{k=1}^\infty \gamma(k) \cos(\omega k) \big]
$$

$$ 
f(\omega) = 1/\pi \big[\beta_1^2 \sigma^2 + \sigma^2 + 2 [\beta_1 \sigma^2] \cos(\omega) \big]
$$

$$ 
f(\omega) = 1/\pi \big[\beta_1^2 \sigma^2 + \sigma^2 + 2 \beta_1 \sigma^2 \cos(\omega) \big]
$$

$$ 
f(\omega) = \sigma^2/\pi \big[\beta_1^2 + 1 + 2 \beta_1 \cos(\omega) \big]
$$

### Produce a plot of the spectrum showing the shape for a few values of $\beta$.


```{r}
# MA(1)
x <- w <- rnorm(1024)
# beta = 0.9
for (t in 2:1024) x[t]<- 0.9 * w[t-1] + w[t]

par(mfrow=c(3,1))
plot(as.ts(x))
acf(x)
spectrum(x, span = 51, log = c("no"))

# beta = 0.7
for (t in 2:1024) x[t]<- 0.7 * w[t-1] + w[t]
par(mfrow=c(3,1))
plot(as.ts(x))
acf(x)
spectrum(x, span = 51, log = c("no"))

# beta = 0.5
for (t in 2:1024) x[t]<- 0.5 * w[t-1] + w[t]
layout(1:3)
par(mfrow=c(3,1))
plot(as.ts(x))
acf(x)
spectrum(x, span = 51, log = c("no"))

# beta = 0.3
for (t in 2:1024) x[t]<- 0.3 * w[t-1] + w[t]
par(mfrow=c(3,1))
plot(as.ts(x))
acf(x)
spectrum(x, span = 51, log = c("no"))

# beta = 1
for (t in 2:1024) x[t]<- 1 * w[t-1] + w[t]
par(mfrow=c(3,1))
plot(as.ts(x))
acf(x)
spectrum(x, span = 51, log = c("no"))

# beta = 3
for (t in 2:1024) x[t]<- 3 * w[t-1] + w[t]
par(mfrow=c(3,1))
plot(as.ts(x))
acf(x)
spectrum(x, span = 51, log = c("no"))
```

---

### 2. Show that if $X_t$ and $Y_t$ are independent, stationary processes with power spectral density functions $f_x(\omega)$ and $f_y(\omega)$, then $V_t=X_t+Y_t$ is also stationary with power spectral density $f_v(\omega)=f_x(\omega)+f_y(\omega)$

$$ f(\omega) = 1/\pi \big[\gamma(0) + 2 \Sigma_{k=1}^\infty \gamma(k) \cos(\omega k) \big] $$

$X_t$ and $Y_t$ are independent, so, $$ \sigma_v^2 = \sigma_x^2 + \sigma_y^2 $$

$$ \gamma_v(h) = \Sigma_{j = 1}^{h} \sigma^2_v \cos(\omega_j h)  $$
$$ = \Sigma_{j = 1}^{h} (\sigma_x^2 + \sigma_y^2) \cos(\omega_j h) $$
$$ = \Sigma_{j = 1}^{h} \sigma_x^2 \cos(\omega_j h) +  \Sigma_{j = 1}^{h} \sigma_y^2 \cos(\omega_j h) $$ 
$$ = \gamma_x(h) + \gamma_y(h)$$

$$ f_v(\omega) = 1/\pi \big[\gamma(0) + 2 \Sigma_{k=1}^\infty \gamma_v(k) \cos(\omega k) \big] $$
$$ = 1/\pi \big[(\gamma_x(0) + \gamma_y(0)) + 2 \Sigma_{k=1}^\infty (\gamma_x(k) + \gamma_y(k)) \cos(\omega k) \big] $$
$$ = 1/\pi \big[(\gamma_x(0) + 2 \Sigma_{k=1}^\infty \gamma_x(k) \cos(\omega k) \big] + 1/\pi \big[(\gamma_y(0) + 2 \Sigma_{k=1}^\infty \gamma_y(k) \cos(\omega k) \big] $$
$$ f_x(\omega) + f_y(\omega) $$

Hence,  $f_v(\omega)=f_x(\omega)+f_y(\omega)$.

-------

## Question 2

The data.frame flow_df contains the average monthly river flow $m^3$/sec in the Mckenzie river at Mckenzie Bridge, Oregon. (I got this from http://robjhyndman.com/tsdldata/askew/askew7.dat who quotes the source: Hipel and Mcleod (1994))
The column time contains a simple time index, the number of months since the start of the record. The column date contains a decimal representation of the date, i.e. 1911.750 is October 1911.

### 1. Estimate the spectrum of the logarithm of flow. Make sure you show evidence you experimented with the amount of smoothing, but you need only show your final plot.


```{r,, message=FALSE, warning=FALSE}
library(ggplot2)
load(url("http://stat565.cwick.co.nz/data/flow_df.rda"))
qplot(time, log(flow), data = flow_df, geom = "line")

attach(flow_df)
flow_log <- log(flow)
# flow_log
spectrum(flow_log)
spectrum(flow_log, spans = 15, taper = 0)
# looks good

#sample trials
# spectrum(flow_log, spans = 20, taper = 0)
# spectrum(flow_log, spans = 23, taper = 0)
# spectrum(flow_log, spans = 50, taper = 0)


```

---

### 2. Fit a smooth trend to the logarithm of flow, and estimate the spectrum of the residuals. How does this spectrum differ from the one in part 1.?

```{r, message=FALSE, warning=FALSE}
attach(flow_df)
fit <- loess(flow_log ~ time, method = "loess")

par(mfrow = c(2,1))
flow_log_ts <- ts(flow_log, start = 1, frequency = 1)
plot(flow_log_ts)
lines(fit$fitted)
plot(fit$residuals, type = "l")

par(mfrow = c(2,1))
spectrum(flow_log, spans = 15, taper = 0)
spectrum(fit$residuals, spans = 15, taper = 0)

abline(v = c(1/12, 2/12), col = "#9E0142")
# 
# abline(v = c(1/12, 2/12, 3/12, 4/12), col = "#9E0142")

str(flow_log_ts)

```

The spectrum of the residual differs from the spectrum of the original data (logarithm of flow) in the first left-most frequencies. This is not surprising as the smoothed trend model we fit earlier, is pretty much a horizontal line/curve, and removing the trend would result in a similar residual series with similar frequencies, except for the level shift. This level shift difference between the two spectrums is visible in the two spectrum plots' Y-axis tick marks. 

---

### 3. Fit a harmonic regression to the residuals from 2 using the estimated spectrum to choose the number and frequencies of the periodic components.


```{r}
par(mfrow = c(1,1))



periodic <- function(x, frequency = 1, order = 1){
  do.call(cbind, lapply(1:order, function(ord){
    cbind(cos(2*pi*ord*frequency*x), sin(2*pi*ord*frequency*x))    
  }))
}

fit_harm <- lm(flow_log ~ time + periodic(time, freq = 1/12, order = 2), data = flow_df)

qplot(time, residuals(fit_harm), data = flow_df, geom = "line") + geom_smooth()
source(url("http://stat565.cwick.co.nz/code/get_acf.R"))
examine_corr(residuals(fit_harm))

library(nlme)
fit_gls <- gls(flow_log ~ time + periodic(time, freq = 1/12, order = 2), data = flow_df,
               correlation = corARMA(p = 1, q = 0))
examine_corr(residuals(fit_gls, type = "normalized"))

```

---

### 4. Examine the residuals from the harmonic regression using both the ACF/PACF and periodogram. Is there any evidence of remaining autocorrelation?

```{r}
# examine the residuals
examine_corr(residuals(fit_gls, type = "normalized"))
spectrum(residuals(fit_gls, type = "normalized"), span = 15) # looks good

```

After fitting a linear autoregression model of AR(1) there was a remaining correlation in the residual. Them, I fit GLS model with AR(1) component to remove the correlation. The residual from this gls harmonic regression with an AR(1) component looks good with uncorrelated white noise property. 





