---
title: "Homework-6"
author: "Tadesse Zemicheal"
date: "February 24, 2016"
output: pdf_document
---
```{r, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message = FALSE, 
  warning = FALSE, results = "hide", fig.height = 3, fig.width = 6) 
```

```{r,echo=F,message=FALSE,warning=F}
library(ggplot2)
library(dplyr)
library(tidyr)
library(TSA)
source(url("http://stat565.cwick.co.nz/code/fortify-ts.r"))
source(url("http://stat565.cwick.co.nz/code/get_acf.R")) # my code for examine_corr
big_font <- theme_grey(base_size =  24)
```
## How does price relate to sales?
```{r}
data(bluebirdlite)
log.sales <- bluebirdlite[, 1]
price <- bluebirdlite[, 2]

plot(log.sales,type='l',col='red')
plot(price,type='l',col='red')
```
### Fit an ols model
```{r}
fit_lm <- lm(log.sales ~ price, data = bluebirdlite)
summary(fit_lm)
qplot(log.sales, price, data = bluebirdlite) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, colour = "red") +
  geom_smooth(se = FALSE)  +
  big_font
```
### Fit model with no correlation

```{r}
fit <- arima(log.sales, xreg = price, order = c(0,0,0))
examine_corr(residuals(fit),lag.max=30)
```
From the ACF and PACF graph the model fits good for arima(1,0,1) and it fits better for AR(4,0) as well.
ACF of arima(1,0,1)

```{r}
#it is okay
fit.2 <- arima(log.sales, xreg = price, order = c(1,0,1))
examine_corr(residuals(fit.2),lag.max=30)
#but AR(4,0) fits good 
```
ACF of arima(4,0,0)

```{r}
fit.4 <- arima(log.sales, xreg = price, order = c(4,0,0))
examine_corr(residuals(fit.4),lag.max=30)
qplot(sample = residuals(fit.4))
fit.4
coef(fit.4)["xreg"] 
confint(fit.4)["xreg", ]

# backtransform and use % decrease with price increase of 0.1
(est <- 100*(1 - exp(0.1 * coef(fit.4)["xreg"])))
(ci <- 100*(1 - exp(0.1 * confint(fit.4)["xreg", ])))

```
It is estimated that an increase in price of 10 cents is associated with decrease in median sales of `r round(est,2)` (95% CI `r round(ci[1], 2)` to `{r} round(ci[2], 2)`). 




# Q2 In class we looked a modelling the relationship between mortality, temperature and particulate matter. Repeat the analysis but seasonally difference all three series first. Compare the results.

### Remove seasonality with S=52
Seasonaliy remove series looks like. 

```{r,warning=F,message=F}
library(ggplot2)
library(dplyr)
# install.packages("tidyr")
library(tidyr)
load(url("http://www.stat.pitt.edu/stoffer/tsa3/tsa3.rda"))
source(url("http://stat565.cwick.co.nz/code/fortify-ts.r"))
source(url("http://stat565.cwick.co.nz/code/get_acf.R")) # my code for examine_corr
big_font <- theme_grey(base_size =  24)

mort <- diff(cmort, lag = 52)
temp <- diff(tempr, lag = 52)
#Remove seasonality with S=52
part <- diff(part, lag = 52)

mort <- data.frame(mortality = mort, part = part, temp = temp)
mort$time <- fortify(cmort)$time[53:508]

# Seasonal deferenced plot 
qplot(time, value, data = gather(mort, variable, value, -time), geom = "line") +
  facet_grid(variable ~ ., scale = "free")
#Checking the model 
```

```{r, fig.show="hold", fig.width=3, fig.height = 2}
qplot(temp, mortality, data = mort)
qplot(part, mortality, data = mort)
qplot(temp, part, data = mort)
```
Relationship between temp and mortality 

```{r}
qplot(temp, mortality, data = mort) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, colour = "red") +
  geom_smooth(se = FALSE)  +
  big_font

qplot(part, mortality, data = mort) + geom_smooth(se = FALSE) + 
  big_font

mort <- mutate(mort, temp_sc = temp - mean(temp),
               temp_2 =temp_sc^2,
               time0 = time - min(time))

fit_lm <- lm(mortality ~ time0 + temp_sc + temp_2 + part, data = mort, na.action = na.omit)
summary(fit_lm)
```

Residuals versus covariates analysis.

```{r}
mort_lm <- fortify(fit_lm)
qplot(time0, .resid, data = mort_lm, geom= "line")
qplot(temp_sc, .resid, data = mort_lm)
qplot(part, .resid, data = mort_lm)
```


###  Assumption and Diagnosis
```{r}
# residuals versus fitted
qplot(.fitted, .resid, data = mort_lm)

# normality of residuals
qqnorm(mort_lm$.resid)
qqline(mort_lm$.resid)
# correlation of residuals
examine_corr(residuals(fit_lm),lag.max=120) 
last_plot() + big_font
# AR (2)? violates regression assumptions
```
ACF and PACF shows there is a significant in the first 2 and at 15.  The best model that fit is AR(15).

```{r}
# two ways to fit
library(nlme)

gls_fit <- gls(mortality ~ time0 + temp_sc + temp_2 + part, data = mort,
    correlation = corARMA(p=15) method = "ML")
summary(gls_fit)

# or
with(mort,auto.arima(mortality,xreg=cbind(time0, temp_sc, temp_2, part)))

arima_fit <- with(mort,arima(mortality, order = c(15, 0, 0),seasonal=list(order=c(2,0,1),period=52),xreg =cbind(time0, temp_sc, temp_2, part)))

arima_fit
examine_corr(arima_fit$residuals,lag.max=110)

auto.arima(mort$mortality,seasonal = T)
fit.5 <- arima(mortality, xreg =cbind(time0, temp_sc, temp_2, part), order = c(2,0,1), 
  seasonal = list(order = c(1, 0, 0), period = 52),data=mort)



```
Only the centered temp is important response variable.

### Diagnosis of the residual.
```{r}
mort$residuals <- residuals(gls_fit, type = "normalized")
mort$fitted <- fitted(gls_fit)
examine_corr(mort$residual,lag.max=105)
```
Looks to fit good. The residual are now white noise. 
```{r, fig.show="hold", fig.width=3, fig.height = 2}
qplot(fitted, residuals, data = mort) + geom_hline(yintercept = 0)
qplot(time, residuals, data = mort) + geom_hline(yintercept = 0)
qplot(temp, residuals, data = mort) + geom_hline(yintercept = 0)
qplot(part, residuals, data = mort) + geom_hline(yintercept = 0)
```

### Forecast temp versus mortality
``` {r}
qqnorm(mort$residuals)
qqline(mort$residuals)
# plot prediction on temp versus mortality plot
mort$pred <- predict(gls_fit, newdata = 
  data.frame(time0 = 0, temp_sc = mort$temp_sc,
             temp_2 = mort$temp_2, part = mean(mort$part)))

qplot(temp, mortality, data = mort) +
  geom_line(aes(y = pred))

```
The analysis shows that removing seasonality by differencing with lag 52, results better fit for the model . Differencing get rid of the main correlation in the noise caused by time.


## Appendix


```{r, all-code, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```





